\documentclass[article,shortnames]{jss}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{amssymb}
%% almost as usual
\author{Stefan R\"odiger$^{\ddag,\S}$\\Lausitz University of\\Applied Sciences\\ \& Charit\'e
	\And Thomas Friedrichsmeier$^{\ddag,\S}$\\Ruhr-University Bochum
	\AND Prasenjit Kapat\\The Ohio State University
	\And Meik Michalke\\Heinrich Heine University\\D\"usseldorf
}
\title{\pkg{RKWard} -- A Comprehensive Graphical User Interface and Integrated Development Environment for Statistical Analysis with \proglang{R}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Stefan R\"odiger, Thomas Friedrichsmeier, Prasenjit Kapat, Meik Michalke} %% comma-separated
\Plaintitle{RKWard -- A Comprehensive Graphical User Interface and Integrated Development Environment for Statistical Analysis with {R}} %% without formatting
\Shorttitle{\pkg{RKWard} a GUI for \proglang{R}} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
  \proglang{R} is a free open-source implementation of the \proglang{S} statistical computing
language and programming environment. The current status of \proglang{R} is a
command line driven interface with no advanced cross-platform graphical user
interface (GUI), but it includes tools for building such. Over the past
years, proprietary and non-proprietary GUI solutions have emerged, 
based on internal or external tool kits, with different scopes and technological concepts.
For example, \proglang{Rgui.exe} and \proglang{Rgui.app} have
become the de facto GUI on the Microsoft Windows and Mac OS X platforms, respectively, for most users. 
In this paper we discuss \pkg{RKWard} which aims to be both a
comprehensive GUI and an integrated development environment 
for \proglang{R}. \pkg{RKWard} is based on the \pkg{KDE} software libraries. Statistical
procedures and plots are implemented using an extendable plugin
architecture based on \proglang{ECMAScript} (\proglang{JavaScript}), \proglang{R}, and XML. \pkg{RKWard}
provides an excellent tool to manage different types of data objects;
even allowing for seamless editing of certain types. The objective of
\pkg{RKWard} is to provide a portable and extensible \proglang{R} interface for both
basic and advanced statistical and graphical analysis, while not
compromising on flexibility and modularity of the \proglang{R} programming
environment itself.

\line(1,0){40}
\item$^{\ddag}$Equal contribution, $^{\S}$Corresponding authors
%%
}
\Keywords{GUI, integrated development environment, plugin, \proglang{R}}

\Plainkeywords{GUI, integrated development environment, plugin, R} %% without formatting
%% at \leftarrowst one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{13}
%% \Issue{9}
%% \Month{September}
%% \Year{2004}
%% \Submitdate{2004-09-29}
%% \Acceptdate{2004-09-29}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Stefan R\"odiger\\
  Lausitz University of Applied Sciences\\
  Department of Bio-, Chemistry and Process Engineering\\
  AND\\
  Center for Cardiovascular Research (CCR)\\
  Charit\'e, Germany\\
  E-mail: \email{stefan\_roediger@gmx.de}\\
  E-mail: \email{rkward-devel@lists.sourceforge.net}\\
}

% \Address{
%  Prasenjit Kapat\\
%  Affiliation\\
%  Department\\
%  E-mail: \email{noname@here.org}
% }
% \Address{
%  Meik Michalke\\
%  Affiliation\\
%  Department\\
% }


%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\input{background}
\input{installing_starting_RKWard}
\input{GUI_elements}
\input{example_session}
\input{conclusion_summary_acknowledgments}

\bibliography{sources}
\end{document}
